(Deep) Induction Rules for GADTs\vspace*{-0.1in}

### 2021-08-18 induction-paper/revised-induction-paper.tex
9-18 revisions on mfps submission

It is not possible to extend structural induction to {\em truly}
nested types, i.e., to nested types whose recursive occurrences appear
below themselves. The quintessential example of such a type is that of
bushes\footnote{To define truly nested types in Agda we must use the
  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define functions over
  them we must use the $\mathsf{TERMINATING}$ flag.  (Similar flags
  are required in Coq.) Although as programmers we know from the
  metatheory in~\cite{jp19} that $\mathsf{Bush}$ is well-defined and
  the functions we define over them terminate, the flags are necessary
  because Agda fails to infer these facts. Analogous comments apply at
  several places below.}\cite{bm98}:

@@

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, each of $\mathsf{pleaf}$ and
$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
regardless of the instances of $\mathsf{PTree}$ appearing in the types
of its arguments. GADTs relax this restriction, allowing their data
constructors both to take as arguments \emph{and return as results}
data whose types involve instances other than the one being
defined. That is, GADTs' constructors' return type instances can, like
that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.\looseness=-1
%these instances can be structured.
%
%Applications of GADTs include generic programming, modeling
%programming languages via higher-order abstract syntax, maintaining
%invariants in data structures, and expressing constraints in embedded
%domain-specific languages. GADTs have also been used, e.g., to
%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

@@

\noindent
where $\mathsf{Q_A : A \to Set}$, $\ol{\mathsf{Q_B : B \to Set}}$,
$\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x : F\,G\,\ol{B}}$.
If we have predicate liftings $\mathsf{F^{\wedge}} : \mathsf{\forall
  (G : Set^{\alpha} \to Set) (\ol{B : Set}) \to (\forall (A : Set) \to
  (A \to Set) \to G\,A \to Set) \to (\ol{B \to Set}) \to F\,G\,\ol{B}
  \to Set}$ for $\mathsf{F}$ and $\mathsf{K^{\wedge}} :
\mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to K\,\ol{B} \to
  Set}$
%\[\begin{array}{lll}
%\mathsf{F^{\wedge}} & : & \mathsf{\forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
%\to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
%\to (\ol{B \to Set}) \to F\,G\,\ol{B} \to Set}\\ 
%\mathsf{K^{\wedge}} & : & \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to
%  K\,\ol{B} \to Set}
%\end{array}\]
for $\mathsf{K}$, then the induction hypothesis $\mathsf{dIndC}$
associated with $\mathsf{c}$ is
\[\begin{array}{l}
\mathsf{dIndC = \lambda (P : \forall (A : Set) \to (A \to Set) \to
  G\,A \to Set) \to}\\
\quad \mathsf{\forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to Set)\,
  (\ol{Q_B : B \to Set})\, (e : Equal\,A\,(K\,\ol{B}))\, (x :
  F\,G\,\ol{B}) \to} \\ 
\quad \mathsf{
  Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
  \to F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x \to
  P\,A\,Q_A\,(c\,\ol{B}\,e\,x)} 
\end{array}\]
and the induction rule for $\mathsf{G}$ is
\[
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to
  Set) \to dIndC\,P \to \forall (A : Set)(Q_A : A \to Set)(y : G\,A)
\to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y} \]

@@

@@

In Sections~\ref{sec:deep-ind-GADTs} and~\ref{sec:framework} we
derived deep induction rules for GADTs that are not truly
nested. Since both nested types and GADTs without true nesting admit
deep induction rules, we might expect truly nested GADTs to admit them
as well. Unfortunately, however, the techniques developed in the
previous sections do not extend to GADTs. Indeed, the induction rule
for a data type generally relies on (unary) parametricity of the model
interpreting it, and deep induction for a truly nested type or a truly
nested GADT also relies on this interpretation being functorial.
Whereas ADTs and nested types both admit functorial parametric
semantics, proper GADTs admit parametric semantics but do not admit
functorial semantics. In this section we show how the techniques
developed in this paper for deriving deep induction rules go wrong for
truly nested GADTs by analyzing the following very simple
example:\looseness=-1

>>>>

It is not possible to extend structural induction to {\em truly
nested types}, i.e., to nested types whose recursive occurrences appear
below themselves. The quintessential example of such a type is that of
bushes\footnote{To define truly nested types in Agda we must use the
  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define functions over
  them we must use the $\mathsf{TERMINATING}$ flag.  (Similar flags
  are required in Coq.) Although as programmers we know from the
  metatheory in~\cite{jp19} that $\mathsf{Bush}$ is well-defined and
  the functions we define over them terminate, the flags are necessary
  because Agda fails to infer these facts. Analogous comments apply at
  several places below.}\cite{bm98}:

@@

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, each of $\mathsf{pleaf}$ and
$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
regardless of the instances of $\mathsf{PTree}$ appearing in the types
of its arguments. GADTs relax this restriction, allowing their data
constructors both to take as arguments \emph{and return as results}
data whose types involve instances other than the one being
defined. That is, GADTs' constructors' return type instances can, like
that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.  For every
GADT in this paper, we require the instance of the return type for
each of its data constructors to be a polynomial in that constructor's
type arguments.\looseness=-1
%these instances can be structured.
%
%Applications of GADTs include generic programming, modeling
%programming languages via higher-order abstract syntax, maintaining
%invariants in data structures, and expressing constraints in embedded
%domain-specific languages. GADTs have also been used, e.g., to
%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

@@

\noindent
where $\mathsf{Q_A : A \to Set}$, $\ol{\mathsf{Q_B : B \to Set}}$,
$\mathsf{e : Equal\,A\,(K\,\ol{B})}$, and $\mathsf{x : F\,G\,\ol{B}}$.
If we have predicate liftings $\mathsf{F^{\wedge}} : \mathsf{\forall
  (G : Set^{\alpha} \to Set) (\ol{B : Set}) \to (\forall (A : Set) \to
  (A \to Set) \to G\,A \to Set) \to (\ol{B \to Set}) \to F\,G\,\ol{B}
  \to Set}$ for $\mathsf{F}$ and $\mathsf{K^{\wedge}} :
\mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to K\,\ol{B} \to
  Set}$
%\[\begin{array}{lll}
%\mathsf{F^{\wedge}} & : & \mathsf{\forall (G : Set^{\alpha} \to Set) (\ol{B : Set})
%\to (\forall (A : Set) \to (A \to Set) \to G\,A \to Set)
%\to (\ol{B \to Set}) \to F\,G\,\ol{B} \to Set}\\ 
%\mathsf{K^{\wedge}} & : & \mathsf{\forall (\ol{B : Set}) \to (\ol{B \to Set}) \to
%  K\,\ol{B} \to Set}
%\end{array}\]
for $\mathsf{K}$, then the induction hypothesis $\mathsf{dIndC}$
associated with $\mathsf{c}$ is
\[\begin{array}{l}
\mathsf{dIndC = \lambda (P : \forall (A : Set) \to (A \to Set) \to
  G\,A \to Set) \to}\\
\quad \mathsf{\forall (A : Set)\, (\ol{B : Set})\, (Q_A : A \to Set)\,
  (\ol{Q_B : B \to Set})\, (e : Equal\,A\,(K\,\ol{B}))\, (x :
  F\,G\,\ol{B}) \to} \\ 
\quad \mathsf{
  Equal^{\wedge}\,A\,(K\,\ol{B})\,Q_A\,(K^{\wedge}\,\ol{B}\,\ol{Q_B})\,e
  \to F^{\wedge}\,G\,\ol{B}\,P\,\ol{Q_B}\,x \to
  P\,A\,Q_A\,(c\,\ol{B}\,e\,x)} 
\end{array}\]
and the induction rule for $\mathsf{G}$ is
\begin{equation}\label{eq:gen-ind-rule}
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to G\,A \to
  Set) \to dIndC\,P \to \forall (A : Set)(Q_A : A \to Set)(y : G\,A)
  \to G^{\wedge}\,A\,Q_A\,y \to P\,A\,Q_A\,y}
\end{equation}

@@

Observing that the above development essentially uses the equality
GADT and its predicate lifting in the discrete category of types to
extend the lifting in~\cite{jp20} --- now specialized to the same
category --- to GADTs, we have established the following theorem:

\begin{theorem}
A GADT $\mathsf{G}$ of the form in \eqref{eq:gadts} admits the deep
induction rule in~\eqref{eq:gen-ind-rule}.
\end{theorem}

@@

{\em Truly nested GADTs} are proper GADTs whose recursive occurrences
appear below themselves. In Sections~\ref{sec:deep-ind-GADTs}
and~\ref{sec:framework} we derived deep induction rules for GADTs that
are not truly nested GADTs.\footnote{Note carefully the distinction
  between a GADT that is not a nested type --- i.e., a proper GADT ---
  and a proper GADT that is not a truly nested GADT.  In fact, truly
  nested types are not proper GADTs and truly nested GADTs are not
  (truly) nested types. Confusion is understandable in light of the
  original, and now well-established, use of the term ``nested type''
  to refer to types that do not necessarily involve actual
  nesting~\cite{bm98}.}  Since both nested types and GADTs without
true nesting admit deep induction rules, we might expect truly nested
GADTs to admit them as well. Unfortunately, however, the techniques
developed in the previous sections do not extend to GADTs. Indeed, the
induction rule for a data type generally relies on (unary)
parametricity of the model interpreting it, and deep induction for a
truly nested type or a truly nested GADT also relies on this
interpretation being functorial.  Whereas ADTs and nested types both
admit functorial parametric semantics, proper GADTs admit parametric
semantics but do not admit functorial semantics. In this section we
show how the techniques developed in this paper for deriving deep
induction rules go wrong for truly nested GADTs by analyzing the
following very simple example:\looseness=-1
